NaN gates and flip FLOPS

### 2019-02-22 nand/paper.tex
Search for impl of nand on binary3

\section{Real numbers}
The real numbers are described by IEEE 754, most recently revised in
2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
exponent. Actually, this understates the elegance of real numbers,
since there are a number of numbers, such as \nan\ (``not a number'')
which are not of this form; \nan\ nas no sign nor mantissa nor
exponent. We also have \inf\ and $-\inf$, which do have a sign, but no
mantissa nor exponent. These are the infinite numbers that you get if
you count very high or very low. Excitingly, we also have both
positive and negative versions of 0.

@@

\nocite{fluint8}

@@

IEEE 754 natively defines several bit widths for floating-point
values, such as the 32-bit binary32 (aka ``single-precision float'')
and 64-bit binary64 (aka ``double-precision float''). The
specification is parameterized to allow other bit widths; for example,
half-precision 16-bit floats are common in GPU code for machine
learning applications (XXX cite). Smaller floats sacrifice precision,
but require less space and allow faster calculations. For our purposes
in this paper, since we only need to represent the two values \nan\ and
\inf, we are interested in the smallest possible representation.

@@

We have one sign bit, two exponent bits, and one mantissa bit, for a
total of four. Since ``single precision'' is 32 bits, ``half
precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
all this works out, shouldn't there be a {\tt eighth} base type in
most modern programming languages and GPUs? Since there are so few
values representable, it would be practical for all the standard
operations to be done in constant time via table lookups. All 16
possible values are given in Figure~\ref{fig:binary4}.

@@

>>>>

\section{Real numbers}
The real numbers are described by IEEE 754, most recently revised in
AD 2008\cite{ieee754}. Every real number has a sign, a mantissa, and an
exponent. Actually, this understates the elegance of real numbers,
since there are a number of numbers, such as \nan\ (``not a number'')
which are not of this form; \nan\ nas no sign nor mantissa nor
exponent. We also have \inf\ and $-\inf$, which do have a sign, but no
mantissa nor exponent. These are the infinite numbers that you get if
you count very high or very low. Excitingly, we also have both
positive and negative versions of 0. Some numbers have multiple
representations, and most numbers cannot be represented at all.

@@

@@

IEEE 754 natively defines several bit widths for floating-point
values, such as the 32-bit binary32 (aka ``single-precision float'')
and 64-bit binary64 (aka ``double-precision float''). The
specification is parameterized to allow other bit widths; for example,
half-precision 16-bit floats are common in GPU code for machine
learning applications (XXX cite). Smaller floats sacrifice precision,
but require less space and allow faster calculations. For our purposes
in this paper, since we only need to represent the two numbers \nan\ and
\inf, we are interested in the smallest possible representation.

@@

We have one sign bit, two exponent bits, and one mantissa bit, for a
total of four. Since ``single precision'' is 32 bits, ``half
precision'' is 16, 4 bits is ``eighth precision.'' Given how nicely
all this works out, shouldn't there be an {\tt eighth} base type in
most modern programming languages and GPUs? Since there are so few
values representable, it would be practical for all the standard
operations to be done in constant time via table lookups. All 16
possible values are given in Figure~\ref{fig:binary4}.

@@

% hardware notes:
% http://openocd.org/ looks like a good way to do in-device
% programming like over SWD, because it runs on the pi. A board
% could have jumpers or even software route to different
% gpio pins. (You would need to run at 3v3 for this of course.)
% I guess make sure the chip is compatible, but there are lots
% explicitly listed.

% Some supported devices:
% AT91SAM7 - no fp in the M7
% at91sam3u - no fp here either :/

% LPC1700, LPC1800,
% LPC2000, LPC4300, , STR7x, STR9x, LM3, STM32x and
% EFM32

% ok the STM32f3 is supported, has FPU, 51 IO pins,
% here is an example breakout:
% http://ebrombaugh.studionebula.com/embedded/stm32f303breakout/index.html
% requires a crystal, voltage regulator, and some filter capacitors,
% but looks manageable. (And the chip has some 5V tolerant TTL I/O
% pins too, which could be nice for other projects)
% also maybe a ferrite bead (for input to analog VDDA, so probably
% not needed for this project)

Idea: Make a FPU. How small could it be?
I guess the simplest thing that could be described as faithful
would compute the NAND function of two binary3 numbers, but on
the entire domain. There are six input pins, three output pins,
and a table size of 8^2 = 64.

A nice simple nand function is
f(x, y) = inf - max(x + y, -inf)

What is a simple implementation of this function?
It can only output inf or nan, so that's good. Of the three
output pins, the first is always 0, the second always 1, so
at worst we would just need a table of 64 bits for this last
value.

Further simplification:

if isfinite(x) && isfinite(y)
then
  // inf - (x + y) = inf
  0
else
// x + y is nan when x is nan, y is nan, or x and y are
// infinites with different signs. If they are both -inf,
// then we have max(-inf, -inf) anyway, which is the same
// as max(nan, -inf). So we have
  if x == inf && y == inf
    // both positive infinities
    // inf - inf = nan
    1
  else
    // inf - -inf = inf
    0

So ultimately this function only returns 1 in the case
that both inputs are exactly +inf, the pattern 0 1 0.

If the inputs are a0 a1 a2, b0, b1, b2, and outputs
are c0, c1, c2, then:
c0 = 0
c1 = 1
c2 = !a0 && !a2 && !b0 && !b2 && a1 && b1

// Could adapt find.sml to find solutions for this,
// using only NAND. Only 2^6 = 64 possible values
// for these variables

let x = nand(a0, a2)   // !a0 || !a2
let y = nand(b0, b2)   // !b0 || !b2

let out =

#  0 & 0 & 0 &   $+0$ \\
#  0 & 0 & 1 &   +1
#  0 & 1 & 0 &   $+\infty$ \\
#  0 & 1 & 1 &    \nan \\
#  1 & 0 & 0 &   $-0$ \\
#  1 & 0 & 1 &   $-1$ \\
#  1 & 1 & 0 &   $-\infty$ \\
#  1 & 1 & 1 &    \nan \\

% would be cool if the meta NAND gate were addressable
% from the rasberry pi (well duh of course it is?) because
% then we could also use it to simulate itself in software
